Dynamical Properties of Molecular Cloud Complexes in Galaxies at the Epoch of Reionization

### 2019-03-05 WRITEUP/ms.tex
edited up to sec 5.2

\begin{abstract}
We study the dynamical properties of molecular clouds complexes (MC) in a prototypical galaxy (\flower)
and its satellite galaxies at the Epoch of Reionization (EoR) using cosmological zoom-in simulation.
We identify MCs using H$_2$ density, 
and compare their masses, sizes, velocity dispersions, gas surface densities, and virial parameters ($\alpha_{\rm vir}$)
with those observed in nearby and \z$\sim$\,2 galaxies.
The MCs span a mass range of $10^{6.5-9}$\,\Msun. Their velocity dispersions and gas surface densities are 
systematically higher than those observed in nearby galaxies, but comparable to starburst galaxies.
We compute the effective two-component Toomre Q parameter (gas and stars) to examine 
regions that are unstable to local axisymmetric perturbations. 
We further perform virial analysis, which is accessible from observations, to examine regions likely to be undergo collapse.
Our analysis indicates that the MCs in the main disk of \flower are unbound, but its 
substructures have lower virial parameters, which is
consistent with the notion that collapsing structures result from gravitational instability within globally stable structures.
% , which are supported by turbulence and rotation on large scale.
On the other hand, we find $\alpha_{\rm vir}\approx$\,2 for the MCs in the satellite galaxies,
which we interpret as result of the weaker stellar feedback and less frequent \SF compared to \flower.
Our results suggest that the more massive and bigger MCs compared to the Milky Way
manifest from the higher gas mass fraction, surface density, and velocity dispersion.
% , which together sets the scale for fragmentation.
High resolution imaging of the first galaxies with the Atacama Large (sub-)Millimeter Array (ALMA) and the
Next Generation Very Large Array (ngVLA)
can furnish scrutinizing data to test our findings and shed light on \SF since the cosmic dark ages.
\end{abstract}
\keywords{methods: data analysis --
          galaxies: high-redshift --
          galaxies: ISM --
          galaxies: evolution --
          galaxies: formation --
          galaxies: starburst --
          stars: formation}

@@

\begin{figure*}[!htbp]
\centering
\includegraphics[trim=10 0 0 0, clip, width=0.65\textwidth]{\figpath/sfh.pdf}
\caption{
    Star formation history of \flower (top right) and
    projected stellar mass distribution of \flower during one of its accretion phases
    at its early stage of evolution {\it (a)}; during one of its major starburst phases
    after a major merger {\it (b)}; and in a relatively quiescent phase post-starburst {\it (c)}.
% f_gas history?
\label{fig:SFH}}
\end{figure*}

@@

We use the virial parameter, which is defined using \Eq{alpha} and describes
the balance between kinetic and gravitational potential energies,
to quantify how stable an MC is.
An $\alpha_{\rm vir}\approx$\,2 corresponds to approximately equipartition between the two energy terms
and is often used to assess the boundedness of a given structure \citep[see e.g., ][]{Kauffmann17b}\footnote{Note, however, that
the true virial state/boundedness of a given structure also depends on the surface terms and the magnetic field (see \Eq{virial}).}.

\subsection{Gravitational Instabilities}   \label{sec:Q}
The onset of gravitational instability is considered to be tightly connected to \SF \citep[e.g.,][]{Kennicutt89a, Wang94a, Li05b, Li06a}.
From virial theorem, the Jeans 
instability sets the scales at which gravitational potential overcomes kinetic energy. Namely, gravitational instabilities 
dominates at scales $L > L_J$, where $L_J$ is the Jeans length. 
However, in a differentially rotating system, centrifugal force resulting from differential rotation can 
inhibit collapse even on these scales. Thus, a disk is unstable in the regime of $L_J < L < L_{\rm rot}$, where $L_{\rm rot}$ is set 
% \propto \Sigma/B^2
by the surface density of the disk and the angular velocity of the rotation.

In theoretical and observational works, gravitational instability has been widely discussed in the 
context of the Toomre stability criterion \citep{Toomre64a, Goldreich65b}, or the Toomre Q parameter 
(or the proxy, circular velocity to dispersion ratio; $v_{\rm circ}/\sigma$; e.g., \citealt{GarciaBurillo03a, Genzel11a, Kassin12a, Leung19a}).
% https://ned.ipac.caltech.edu/level5/March15/Glazebrook/Glazebrook5.html
%
% from linear analysis
For {\it axisymmetric} mode ($m$\eq0), the dispersion relation for the growth of density perturbation in a rotating turbulent disk
of finite thickness $h$ 
is described by:
\begin{equation}
\omega^2 = \kappa^2 - \frac{2\pi G \Sigma |k|}{1 + |k| h} + \sigma^2 k^2,
\label{eqn:3Ddisp}
\end{equation}
where $k$ is the wavenumber and $\kappa$ is the epicyclic frequency, defined as:
\begin{equation}
\kappa^2\equiv\frac{2\Omega}{R}\frac{d}{dR}\left(R^2\Omega\right).
\label{eqn:kappa}
\end{equation}
% \kappa\propto\Omega
The first term on the RHS of Equation~\ref{eqn:3Ddisp} describes rotation support, whereas the second term describes self-gravity, and
the third term describes pressure support.
One can then define the so-called Toomre-Q parameter as:
\begin{equation}
Q\equiv\frac{\sigma\kappa}{\pi G \Sigma}.
\label{eqn:Q}
\end{equation}
For $kh\ll1$, Equation~\ref{eqn:3Ddisp} reduces to the thin disk approximation, which 
becomes unstable % perturbation grows exponentially in time
for some $k$ when $\omega^2 < 0$ (or equivalently, $Q < 1$; \citealt{Toomre64a}).
Under this condition (i.e., when local gravitational instability 
overcomes support from pressure and rotation), fragmentation {\it can} happen. 
% but further evolution depends on their EOS.
%(e.g., instability impose by angular momentum from gas accretion from the cosmic web and
%satellite galaxies and stellar feedback).
Similarly, for a collisionless fluid, the Toomre Q parameter is defined by:
\begin{equation}
Q\equiv\frac{\sigma\kappa}{3.36 G \Sigma}
\end{equation}
\citep{Toomre64a}.

To account for stabilizing effect of both gas and stars and the non-negligible disk thickness, 
% since thickness of a disk weakens the gravitational forces within the disk plane and lower the values for $Q$
% Romeo:  The effect of thickness is to increase the stability parameter of each 
% component by a factor T, which depends on the ratio of vertical to radial velocity dispersion, following eqn 8 of their paper.
we consider the approximation for the effective two-component $Q_{\rm eff}$ parameter following \citet{Romeo11a, Inoue16a}
which is expressed as follows:
\begin{equation}
Q^{-1}_{\rm eff} = \[ \left\{
				\begin{array}{lccr}
					     {\displaystyle\frac{w}{T_{\rm star} Q_{\rm star}} + \frac{1}{T_{\rm gas} Q_{\rm gas}}}      & & & \mbox{if\ }T_{\rm star} Q_{\rm star} \geq T_{\rm gas} Q_{\rm gas} \\ [0.75em]
                                               {\displaystyle\frac{1}{T_{\rm star} Q_{\rm star}} + \frac{w}{T_{\rm gas} Q_{\rm gas}}}      & & & \mbox{if\ }T_{\rm star} Q_{\rm star} \leq T_{\rm gas} Q_{\rm gas} \\				
				\end{array} 
			    \right,
			    \]
\end{equation}
where $w$ is defined as:
\begin{equation}
w\equiv\frac{2 \sigma_{\rm star} \sigma_{\rm gas}}{\sigma_{\rm star}^2 + \sigma_{\rm gas}^2},
\end{equation}
and $T$ is approximated as:
\begin{equation}
T = \[ \left\{
		\begin{array}{lccr}
			{\displaystyle 0.8 + 0.7\left(\frac{\sigma_{z}}{\sigma_{r}}\right)}      && & \mbox{if\ } \sigma_z \lesssim 0.5 \times \sigma_r \\ [0.75em]
			{\dispalystyle 1 + 0.6 \left(\frac{\sigma_z}{\sigam_r}\right)}             & & & \mbox{if\ } \sigma_z \gtrsim 0.5 \times \sigma_r 
\\
		\end{array}
		\right.
		\]
\end{equation}

@@

% Molecular clouds are the natal place for \SF, their structure and dynamics hold important clues to understanding
% the mechanisms and physics of formation and evolution of molecular structures and \SF.
In next section, we examine the MC properties in the context of the Larson's relations \citep{Larson81a}, which
describe interactions between gravity and turbulence and is the first set of
relations used for studying star-forming molecular structure formation based on observables (linewidth-size,
density-size, and mass-size relations) and for comparing properties of molecular structures
in different galactic environments.
% Inner MW: R=2-8.5 kpc, most of the mass in the molecular ISM is in the form of GMCs with sizes $gtrsim$20\,pc and masses greater than 10$^5$\,\Msun \citep{Sanders85a}.

@@

The mass of an MC is calculated from the uniformly-gridded 3D density field,
integrating over all zones of the MC, multiplied by its volume.
The size of an MC ($R$) is defined assuming spherical geometry (i.e.,
the size is parameterized via the radius of a sphere, which has a volume corresponding to that of the identified MC).
% {\AP The size is an effective radius, calculated from the total volume of the MC assuming spherical symmetry.}
Since in \obs, dense gas contributes more to the observed linewidths than the more diffuse gas. We therefore calculate
the non-thermal component of the velocity dispersion ($\sigma_{\rm NT}$) of each MC as a density-weighted quantity:
\begin{equation}
\sigma_{\rm NT}^2 = \frac{1}{{3}}\frac{\sum_{i} \rho_i \left|\mathbf{v}_i - \mathbf{\bar v}\right|^2}{\sum_i \rho_i}\,,
\end{equation}
which is summed over the cells composing each MC.
The total velocity dispersion includes contributions from the thermal sound speed ($c_s$):
\begin{equation}
\sigma^2 = \sigma_{\rm NT}^2 + c_s^2.
\end{equation}
The local sound speed ($c_s$)
of all identified molecular complexes are generally much smaller compared to the their
turbulent velocities\footnote{We find comparable velocity dispersion between that derived from
taking the root mean square versus that derived from thermal and non-thermal pressure terms, indicating that
rotation velocity is unlikely to be the dominant source of velocity dispersion reported here.
% We further quantify the contributions arising from galactic rotation and shear to the extracted velocity dispersions of the MCs in \Sec{Q}.
In the subsequent sections, we adopt the root mean square of the velocity field
as the velocity dispersion ($\sigma$).}.
On average, the Mach number of the MCs corresponds to about $\mathcal{M}\approx50$.
This is consistent with the global analysis done on \flower by \citet{Vallini18a}.

@@

Maps of the Toomre Q parameter of the gas component are shown in \Fig{Q} for 
different projection angles during the accretion phase of \flower (see \Fig{h2density} for examples of clumps identified).

\begin{figure*}[htbp]
\centering
\includegraphics[trim=0 20 30 30, clip, width=0.85\textwidth]{\figpath/ss_16_toomre_proj_0}
\caption{
Maps of the total gas surface density (top left), 
velocity dispersion (top right),
epicyclic frequency (bottom left), 
and Toomre Q parameter (bottom right) in the $xy$-plane.
Modest smoothing has been applied to the maps. A divergent colormap is used for the Toomre Q map to facilitate 
identification of regions above and below $\log{Q_{\rm gas}}$\eq0.
\label{fig:Q}}
\addtocounter{figure}{-1}
\end{figure*}

@@

For MCs in the satellite galaxies, we find low virial parameters and pressures, suggesting 
that these structures are likely collapsing structures. 
Their virial parameters are comparable with those found in clouds that are formed in the galactic disks 
($\alpha_{\rm vir}$\eq0.2$-$10; \citealt{Dobbs08a, Tasker09a}).
Thus, \SF continues as gas from the satellite galaxies is being accreted onto the main galaxies during the EoR.

@@

\subsection{Origin of the MCs} \label{sec:origin} 
\DL{discuss stuff on Toomre Q maps with clumps identified.}

>>>>

\begin{abstract}
We study the dynamical properties of molecular clouds complexes (MC) in a prototypical galaxy (\flower)
and its satellite galaxies at the Epoch of Reionization (EoR) using cosmological zoom-in simulation.
We identify MCs using H$_2$ density, 
and compare their masses, sizes, velocity dispersions, gas surface densities, and virial parameters ($\alpha_{\rm vir}$)
with those observed in nearby and \z$\sim$\,2 galaxies.
The MCs span a mass range of $10^{6.5-9}$\,\Msun. Their velocity dispersions and gas surface densities are 
systematically higher than those observed in nearby galaxies, but comparable to starburst galaxies.
We compute the effective two-component Toomre $Q$ parameter (gas and stars) to examine 
regions that are unstable to local axisymmetric perturbations. 
We further perform virial analysis, which is accessible from observations, to examine regions likely to be undergo collapse.
Our analysis indicates that the MCs in the main disk of \flower are unbound, but its 
substructures have lower virial parameters, which is
consistent with the notion that collapsing structures result from gravitational instability within globally stable structures.
% , which are supported by turbulence and rotation on large scale.
On the other hand, we find $\alpha_{\rm vir}\approx$\,2 for the MCs in the satellite galaxies,
which we interpret as result of the weaker stellar feedback and less frequent \SF compared to \flower.
Our results suggest that the more massive and bigger MCs compared to the Milky Way
manifest from the higher gas mass fraction, surface density, and velocity dispersion.
% , which together sets the scale for fragmentation.
High resolution imaging of the first galaxies with the Atacama Large (sub-)Millimeter Array (ALMA) and the
Next Generation Very Large Array (ngVLA)
can furnish scrutinizing data to test our findings and shed light on \SF since the cosmic dark ages.
\end{abstract}
\keywords{methods: data analysis --
          galaxies: high-redshift --
          galaxies: ISM --
          galaxies: evolution --
          galaxies: formation --
          galaxies: starburst --
          stars: formation}

@@

\begin{figure*}[!htbp]
\centering
\includegraphics[trim=10 0 0 0, clip, width=0.65\textwidth]{\figpath/sfh.pdf}
\caption{
    Star formation history of \flower (top right) and
    projected stellar mass distribution of \flower during 
    {\it (a)} one of its accretion phases
    at its early stage of evolution; 
    {\it (b)} one of its major starburst phases
    after a major merger; and
    {\it (c)} a relatively quiescent phase post-starburst.
% f_gas history?
\label{fig:SFH}}
\end{figure*}

@@

We use the virial parameter, which is defined by \Eq{alpha} and describes
the balance between kinetic and gravitational potential energies,
to quantify how stable an MC is.
An $\alpha_{\rm vir}\approx$\,2 corresponds to approximately equipartition between the two energy terms
and is often used to assess the boundedness of a given structure \citep[see e.g., ][]{Kauffmann17b}\footnote{Note, however, that
the true virial state/boundedness of a given structure also depends on the surface terms and the magnetic field (see \Eq{virial}).}.

\subsection{Gravitational Instabilities}   \label{sec:Q}
The onset of gravitational instability is considered to be tightly connected to \SF \citep[e.g.,][]{Kennicutt89a, Wang94a, Li05b, Li06a}.
From virial theorem, Jeans 
instability sets the scales at which gravitational potential overcomes kinetic energy. Namely, gravitational instabilities 
dominates at scales $L > L_J$, where $L_J$ is the Jeans length. 
However, in a differentially rotating system, centrifugal force resulting from differential rotation can 
inhibit collapse even on these scales. Thus, a disk is thought to be 
unstable on scales of $L_J < L < L_{\rm rot}$, where $L_{\rm rot}$ is set 
% \propto \Sigma/B^2
by the surface density of the disk and the angular velocity of the rotation.

In theoretical and observational works, gravitational instability has been widely discussed in the 
context of the Toomre stability criterion \citep{Toomre64a, Goldreich65b}, or the Toomre $Q$ parameter 
(or the proxy, circular velocity to dispersion ratio; $v_{\rm circ}/\sigma$; e.g., \citealt{GarciaBurillo03a, Genzel11a, Kassin12a, Leung19a}).
% https://ned.ipac.caltech.edu/level5/March15/Glazebrook/Glazebrook5.html
%
% from linear analysis
For {\it axisymmetric} mode ($m$\eq0), the dispersion relation for the growth of density perturbation in a rotating turbulent disk
of finite thickness $h$ 
is described by:
\begin{equation}
\omega^2 = \kappa^2 - \frac{2\pi G \Sigma |k|}{1 + |k| h} + \sigma^2 k^2,
\label{eqn:3Ddisp}
\end{equation}
where $k$ is the wavenumber and $\kappa$ is the epicyclic frequency, defined as:
\begin{equation}
\kappa^2\equiv\frac{2\Omega}{R}\frac{d}{dR}\left(R^2\Omega\right).
\label{eqn:kappa}
\end{equation}
% \kappa\propto\Omega
The first term on the RHS of Equation~\ref{eqn:3Ddisp} describes rotation support, whereas the second term describes self-gravity, and
the third term describes pressure support.
One can then define the so-called Toomre-Q parameter as:
\begin{equation}
Q\equiv\frac{\sigma\kappa}{\pi G \Sigma}.
\label{eqn:Q}
\end{equation}
For $kh\ll1$, Equation~\ref{eqn:3Ddisp} reduces to the thin disk approximation, which 
becomes unstable % perturbation grows exponentially in time
for some $k$ when $\omega^2 < 0$ (or equivalently, $Q < 1$; \citealt{Toomre64a}).
Under this condition (i.e., when local gravitational instability 
overcomes support from pressure and rotation), fragmentation {\it can} happen. 
% but further evolution depends on their EOS.
%(e.g., instability impose by angular momentum from gas accretion from the cosmic web and
%satellite galaxies and stellar feedback).
Similarly, for a collisionless fluid, the Toomre $Q$ parameter is defined by:
\begin{equation}
Q\equiv\frac{\sigma\kappa}{3.36 G \Sigma}
\end{equation}
\citep{Toomre64a}.

To account for stabilizing effect of both gas and stars and the non-negligible disk thickness, 
% since thickness of a disk weakens the gravitational forces within the disk plane and lower the values for $Q$
% Romeo:  The effect of thickness is to increase the stability parameter of each 
% component by a factor T, which depends on the ratio of vertical to radial velocity dispersion, following eqn 8 of their paper.
we consider the approximation for the effective two-component $Q_{\rm eff}$ parameter following \citet{Romeo11a} and
\citet{Inoue16a} which is expressed as follows:
\begin{equation}
Q^{-1}_{\rm eff} = \[ \left\{
				\begin{array}{lccr}
					     {\displaystyle\frac{w}{T_{\rm star} Q_{\rm star}} + \frac{1}{T_{\rm gas} Q_{\rm gas}}}      & & & \mbox{if\ }T_{\rm star} Q_{\rm star} \geq T_{\rm gas} Q_{\rm gas} \\ [0.75em]
                                               {\displaystyle\frac{1}{T_{\rm star} Q_{\rm star}} + \frac{w}{T_{\rm gas} Q_{\rm gas}}}      & & & \mbox{if\ }T_{\rm star} Q_{\rm star} \leq T_{\rm gas} Q_{\rm gas} \\				
				\end{array} 
			    \right,
			    \]
\end{equation}
where $w$ is defined as:
\begin{equation}
w\equiv\frac{2 \sigma_{\rm star} \sigma_{\rm gas}}{\sigma_{\rm star}^2 + \sigma_{\rm gas}^2},
\end{equation}
and $T$ is approximated as:
\begin{equation}
T = \[ \left\{
		\begin{array}{lccr}
			{\displaystyle 0.8 + 0.7\left(\frac{\sigma_{z}}{\sigma_{r}}\right)}      && & \mbox{if\ } \sigma_z \lesssim 0.5 \times \sigma_r \\ [1.25em]
			{\displaystyle 1 + 0.6\left(\frac{\sigma_{z}}{\sigma_{r}}\right)}        & & & \mbox{if\ } \sigma_z \gtrsim 0.5 \times \sigma_r 
\\
		\end{array}
	\right \rlap{.}
	\]
\end{equation}

For the purpose of this work, we present Toomre $Q$ map in \Sec{results} and discuss the importance of support against axisymmetric perturbations from galactic rotation and shear in the extracted velocity dispersions of the MCs in \Sec{origin}.
\DL{so does the extracted velocity disp. has a contribution from shear or is it just pressure? I thought we have concluded 
that it is just high pressure leading to the high velocity disp, but not because of diff. rotation. It's not clear to me how 
the Toomre $Q$ map is helpful to inform us on how much of the extracted velo. disp. is coming from diff. rotation.}

@@

% Molecular clouds are the natal place for \SF, their structure and dynamics hold important clues to understanding
% the mechanisms and physics of formation and evolution of molecular structures and \SF.
In next section, we examine and compare properties of the MCs with
observations in the context of the Larson's relations \citep{Larson81a}. 
Namely, the linewidth-size, density-size, and mass-size relations.
Larson's relations are adopted for the discussion in this work as they 
describe the interactions between gravity and turbulence, and is the first set of
relations used for studying star-forming molecular structure formation based on observables
which has been routinely used for comparing properties of molecular structures in different galactic environments.
% Inner MW: R=2-8.5 kpc, most of the mass in the molecular ISM is in the form of GMCs with sizes $gtrsim$20\,pc and masses greater than 10$^5$\,\Msun \citep{Sanders85a}.

@@

The mass of an MC is calculated from the uniformly-gridded 3D density field,
integrating over all zones of the MC, multiplied by its volume.
The size of an MC ($R$) is defined assuming spherical geometry (i.e.,
the size is parameterized via the radius of a sphere, which has a volume corresponding to that of the identified MC).
% {\AP The size is an effective radius, calculated from the total volume of the MC assuming spherical symmetry.}
Since in \obs, dense gas contributes more to the observed linewidths than the more diffuse gas. We therefore calculate
the non-thermal component of the velocity dispersion ($\sigma_{\rm NT}$) of each MC as a density-weighted quantity:
\begin{equation}
\sigma_{\rm NT}^2 = \frac{1}{{3}}\frac{\sum_{i} \rho_i \left|\mathbf{v}_i - \mathbf{\bar v}\right|^2}{\sum_i \rho_i}\,,
\end{equation}
which is summed over the cells composing each MC.
The total velocity dispersion includes contributions from the thermal sound speed ($c_s$):
\begin{equation}
\sigma^2 = \sigma_{\rm NT}^2 + c_s^2.
\end{equation}
The local sound speed of all identified molecular complexes are much smaller compared to the their
turbulent velocities\footnote{We find comparable velocity dispersion between that derived from
taking the root mean square versus that derived from thermal and non-thermal pressure terms, indicating that
rotation velocity is unlikely to be the dominant source of velocity dispersion reported here.
% We further quantify the contributions arising from galactic rotation and shear to the extracted velocity dispersions of the MCs in \Sec{Q}.
In the subsequent sections, we adopt the root mean square of the velocity field
as the velocity dispersion ($\sigma$).}.
On average, the Mach number of the MCs corresponds to about $\mathcal{M}\approx50$.
This is consistent with the global analysis done on \flower by \citet{Vallini18a}.

@@

Maps of the gas Toomre $Q$ parameter and the effective two-component $Q_{\rm eff}$ parameter along different projection angles during the accretion phase of \flower are shown in \Fig{Q} and \Fig{Q_eff}, respectively (see \Fig{h2density} for examples of clumps identified 
in this snapshot).

\begin{figure*}[htbp]
\centering
\includegraphics[trim=0 20 30 30, clip, width=0.85\textwidth]{\figpath/ss_16_toomre_proj_0}
\caption{
Maps of the total gas surface density (top left), 
velocity dispersion (top right),
epicyclic frequency (bottom left), 
and Toomre $Q$ parameter (bottom right) in the $xy$-plane.
Modest smoothing has been applied to the maps. A divergent colormap is used for the Toomre $Q$ map to facilitate 
identification of regions above and below $\log{Q_{\rm gas}}$\eq0.
\label{fig:Q}}
\addtocounter{figure}{-1}
\end{figure*}

@@

For MCs in the satellite galaxies, we find low virial parameters and pressures, suggesting 
that these structures are likely collapsing structures. 
Their virial parameters are comparable with those found in clouds that are formed in the galactic disks 
($\alpha_{\rm vir}$\eq0.2$-$10; \citealt{Dobbs08a, Tasker09a}).
Thus, this would suggest 
\SF continues as gas from the satellite galaxies is being accreted onto the main galaxies during the EoR.

@@

\subsection{Origin of the MCs} \label{sec:origin} 
\DL{discuss stuff on Toomre $Q$ maps with clumps identified.}
